One-stage point transect distance sampling using iterated integrated nested Laplace approximations

### 2024-11-05 paper.tex
final edits

The model fits in roughly 40 to 60 seconds on a laptop with 10 Gb RAM and a 2.50 GHz CPU when INLA is run using the `empirical Bayes' option which does not incorporate uncertainty in hyper parameters.  The model fits in a few minutes in the full INLA mode, including hyper parameter uncertainty.  We suggest using the `empirical Bayes' mode for model development and full INLA on a smaller subset of candidate models.

The predicted mean of the posterior intensity field is shown in \autoref{fig:intensity-mean-cv}A along with a map of the coefficient of variation (CV) (\autoref{fig:intensity-mean-cv}B) and standard deviation of the posterior intensity field (\autoref{fig:intensity-mean-cv}C) on a regular prediction grid with each prediction cell area approximately 1.7 hectares.
\begin{figure}[!htb]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/intensity_mean_cv_sd.pdf}
		\caption{The A) mean, B) CV and C) SD of the posterior intensity field.  Units are per m$^2$.  The colour scale is different for each panel.}
		\label{fig:intensity-mean-cv}
	\end{center}
\end{figure}
This shows a region of high intensity in the south and much lower intensity in the north, agreeing with a standard two-stage analysis of the same data \citep{camp_dsm_2020}.  The CV plot shows that the CV in the posterior intensity field is lower in areas with greater sampling effort and higher intensity.  There is a clear indication of preferential sampling with more survey effort in the south, where densities are higher, compared to the north.  In the north, estimated intensities are lower, which also contributes to larger CV values and makes the CV map hard to interpret.  The standard deviation map shows stronger overall posterior variability in the south where intensity is larger, most likely due to the assumed log-normal relationship between the random field and the Poisson rate parameter.

We also map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction location (Figure \ref{fig:intensity-quantiles}).  As we discuss below, these quantiles are independent summary statistics of the posterior intensity field at each prediction location.  
\begin{figure}[!htb]
	\begin{center}
		\includegraphics{figures/intensity_quantiles.pdf}
		\caption{Pointwise posterior quantiles for the intensity:  A) 0.025 quantile B) 0.975 quantile}
		\label{fig:intensity-quantiles}
	\end{center}
\end{figure}

The posterior detection function (\autoref{fig:dist-mat-pairs}) shows that detectability drops to just under 0.25 at the maximum observable distance of 58 metres and also broadly agrees with a two-stage analysis.
\begin{figure}[!htb]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/dist_mat_pairs.pdf}
		\caption{A) Posterior half-normal detection function. B) Posterior Mat\'ern correlation function.  The black line is the posterior mean and the grey shaded area is 95\% credible interval in both plots. C) Boxplots of the frequency of pairwise distances based on 500 simulated datasets from the posterior predictive distribution.  Axis labels mark the largest value in each 1km bin.   Red circles indicate the observed count of pairwise distances within each bin.}
		\label{fig:dist-mat-pairs}
	\end{center}
\end{figure}
The posterior Mat\'ern correlation function is shown in \autoref{fig:dist-mat-pairs}B; the grey regions show the 95\% credible interval.  We note that this correlation function shows positive correlation even at distances of 10,000 metres, wider than the full length of the study region (approximately 8,000 metres).  This is because the correlation function is defined on the log-intensity scale.  On the intensity scale the empirical range of correlation is shorter.

To evaluate the ability of the SPDE effect to capture the spatial heterogeneity in the observed data we generated 1,000 point patterns using the posterior intensity field and thinned them with the posterior detection function to create 1,000 simulated datasets to compare with the observed data.  We calculated the pairwise distances between all observations for each of these datasets and compared the frequency of pairwise distances with those in the observed data.

Figure  shows that the SPDE effect broadly captures the spatial clustering in the observed data.  There is some evidence of under-predicting the strength of the clustering at shorter distances between 1,000 and 2,500 m, with the simulated data showing some evidence of negative bias.  This could be due to over-smoothing in some areas which can happen if using a stationary model.  Large areas of the study region having low density, which has a larger correlation range than higher density areas, and so the effect will average between these.

Figure \autoref{fig:dist-mat-pairs}C is similar in spirit to Ripley's $K$-function \citep{ripley_SecondorderAnalysisStationary_1976}, which is a common approach to investigating spatial structure in point patterns.  The pairwise distances in the observed data is a function of spatial heterogeneity as well as the spatially restricted sampling scheme.  Other methods of comparison could also be considered such as the average difference between predicted and observed counts at each point transect.  Working with posterior predictive datasets is a flexible approach to model evaluation that allows for innovative methods of comparison with observed data that can focus on particular features of the model.

To plot the posterior abundance estimate for the \akepa{} in Hakalau we use a Monte Carlo method based on sampling realisations of the posterior intensity field.  Let $n$ denote the abundance of the population within the region of interest and $N = \int_{\Omega}\lambda(\bs)\mathrm{d}\bs$ its corresponding random variable within our Bayesian framework.  Integrating the mean of the posterior intensity field provides a point estimate for the abundance, i.e. the expected abundance.  To generate a full posterior estimate for abundance we approximate the posterior $\pi(N | \bm{Y})$ by taking $m$ Monte Carlo samples of the posterior intensity field, $\lambda^{(1)}, \ldots, \lambda^{(m)}$ and approximate the posterior for the abundance as $\pi(N | \bm{Y}) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda^{(i)}, \bm{Y})$. Each $\pi(N | \lambda = \lambda^{(i)}, \bm{Y})$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda^{(i)}(\bs)\mathrm{d}\bs$. 
\begin{figure}[!htb]
	\begin{center}
		\includegraphics{figures/N_posterior.pdf}
		\caption{Posterior abundance.  Grey area marks two Monte Carlo standard errors above and below the mean.}
		\label{fig:realized-abundance-posterior}
	\end{center}
\end{figure}

\autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.  This allows us to estimate the probability for any specific value of realised abundance $n$ which may be of interest to conservation managers.  This approach can also be used to generate credible intervals and other posterior summaries of interest.

The abundance estimates cannot be directly compared to the results in \cite{camp_dsm_2020} since their two-stage model also includes a temporal correlation and we include an area of Hakalau that the authors excluded.  Converting the above abundance estimates to density per hectare for the whole study region allows for some comparison with Figure 2 in \cite{camp_dsm_2020} which suggests that \akepa{} density was roughly between 0.5 and 1.5 birds per hectare in 2002 (considering both the design-based estimate and the spatio-temporal GAM estimate).  In our analysis we find a credible region of roughly 0.9 to 1.7 birds per hectare, which is slightly different but overlaps significantly with the results in \cite{camp_dsm_2020}.

An attractive feature of our Bayesian one-stage model is that, by basing model outputs on posterior Monte Carlo samples, all the outputs presented in the above, the summaries of the random field as well as the posterior abundance, incorporate the uncertainty in the detection model in a natural way.  There is no additional step required to incorporate detection function uncertainty beyond working with samples from the joint posterior.

@@

Our presentation of the results above is broadly consistent with approaches taken in the species distribution modelling literature, although maps of predictive uncertainty (Figures \ref{fig:intensity-mean-cv} and \ref{fig:intensity-quantiles}) are not always provided.  It is common to see maps of point estimates (e.g. a map of expected density) without any accompanying communication of uncertainty.  It is also common to report a point estimate of the expected abundance along with uncertainty in the point estimate.  However, uncertainty in a point estimate will have lower variance than the variance of the posterior random variable.  For this reason we chose to present the full abundance posterior in \autoref{fig:realized-abundance-posterior} as opposed to uncertainty of a `best guess' for abundance.  Our intention in this section is to highlight further limitations with the model outputs presented above and to suggest ways to address these limitations.

The most common method to communicate uncertainty in maps of animal density is to produce maps of some measure of predictive uncertainty, such as the standard deviation or CV \citep{fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014}.  These can be derived from the posterior predictive distribution for the model in a Bayesian framework and bootstrapping in a maximum likelihood context.  Another approach is to map quantiles or the probability of exceeding certain thresholds at each prediction location \citep{russell_avoidance_2016, wilson_hierarchical_2010}.  In Section \ref{sec-results}, we present maps of the CV, standard deviation, and the 0.025 and 0.975 quantiles of the posterior intensity field (Figures \ref{fig:intensity-mean-cv}B, \ref{fig:intensity-mean-cv}C, \ref{fig:intensity-quantiles}A, and \ref{fig:intensity-quantiles}B, respectively) that are all intended to communicate uncertainty in the estimate of the \akepa{} spatial distributions.  Whilst such maps that summarise spatial predictions are useful, they all mask certain properties that may be important when communicating the results of the analysis.

A map showing the posterior predicted mean across the study region (Figure \ref{fig:intensity-mean-cv}A) is often the key output of a species distribution model.  However, the posterior mean will always be smoother than realisations of the field itself, highlighting that even this relatively innocuous summary statistic can mask important features of the random field.  \autoref{fig:intensity-realizations} shows three such realisations from the posterior intensity field.  Note that each realisation has a finer-grained spatial structure than is shown in the posterior mean (\autoref{fig:intensity-mean-cv}A).  Our interpretation is that the clustering of animals is stronger than might be expected if we only looked at the map of the posterior mean.
\begin{figure}[!htb]
	\includegraphics{figures/intensity_realized.pdf}
	\caption{Three realizations of the posterior intensity field}
	\label{fig:intensity-realizations}
\end{figure}

For model evaluation we recommend plotting multiple realisations of the posterior field as these will more closely resemble the spatial structure of the observed data and thus what ecologists might encounter in the field.  The mean can give a sense of homogeneity, particularly in the high density area in the south, when compared to realisations of the mean.  Presenting multiple realisations, perhaps as an animation, is also an effective way to communicate uncertainty. \cite{bowman_GraphicsUncertainty_2019a} provide an interpolation method that preserves the mean and covariance structure of realisations that smoothly interpolates between realisations.  This avoids abrupt changes that can occur when animating the raw realisations.

The CV map (\autoref{fig:intensity-mean-cv}B) is also intended to communicate uncertainty.  However, CV values will generally be higher in regions of low predicted intensity, particularly if the posterior standard deviation is relatively consistent across the study region or positively correlated with the intensity.  Higher CV values in Figure \ref{fig:intensity-mean-cv}B represent areas of lower intensity but not necessarily higher uncertainty.  However, there is also variable sampling effort within the study region, with higher CV values in the region with lower effort.  It is not clear what impact this has on the CV values and this makes the map hard to interpret.

The standard deviation (SD) map (\autoref{fig:intensity-mean-cv}C) shows spatially varying SD values also intended to indicate uncertainty.  However, care should be taken as to what the mapped values imply in the context of a given analysis.  A default colour scale will nearly always show some spatial variation in uncertainty, with some regions of relatively high and low uncertainty.  However, whether these relative differences matter is dependent on the overall context of the analysis.  Some levels of variability may be acceptable in some contexts and unacceptable in others.

The SD values also reflect a modelling assumption rather than variability due to, say, varying survey effort.  The SD is linked to the predicted mean by the assumed log-Gaussian relationship.  If X is normally distributed with mean $\mu$ and variance $\sigma^2$ then $Z = \exp(X)$ has a mean-variance relationship given by $\text{Var}(Z) = \exp(\sigma^2 - 1)\mathbb{E}(Z)^2$. Therefore the SD map will generally show a pattern similar to the mean intensity map (Figures 2A and 2B).  Since this relationship is the result of a modelling assumption and not necessarily reflective of the detectability of animals or survey effort.  It should be noted however, that this modelling assumption affects uncertainty quantification for all outputs based on the posterior intensity field, not just the SD map.

The quantile maps (Figure \ref{fig:intensity-quantiles}) are also difficult to interpret. The temptation is to perceive the maps as showing possible intensity surfaces that could have produced the observed data. One possible intensity surface (0.025 quantile map) that signifies a kind of lower bound on abundance and another surface (0.975 quantile map) that signifies higher abundance.  However, presenting these quantiles together in a single map obscures the fact that it is vanishingly unlikely for the random field to \textit{simultaneously} achieve the 0.025 or 0.975 quantiles at all prediction locations.  A similar problem holds for maps of the marginal probability of exceeding a threshold at each prediction location.

Presenting independent quantiles jointly in a single image is risky.  These caveats make the map difficult to use for most non-statistically trained audiences and even trained statisticians may misinterpret these quantile maps.  To demonstrate the possible consequences of this we (incorrectly) treat the lower and upper quantile plots as though they are intensity functions and integrate them to obtain an expected abundance estimate of approximately 2,900 for the 0.025 quantile map and 10,700 for the 0.975 quantile map.  A naive (and tempting) interpretation of these numbers is as lower and upper limits of the 95\% credible interval for abundance. However, these abundance estimates are not supported by the posterior for abundance (\autoref{fig:realized-abundance-posterior}).  This shows that the tendency to interpret these maps as showing possible intensity surfaces can lead to interpretations that are inconsistent with the very model that generated the maps.

We conclude that, although all useful in their own way, each of the mapped summaries of the posterior intensity field suffer from some weaknesses, whether through masking certain properties of the random field or through being difficult or impossible to interpret. In the following section we suggest an approach that avoids some, but not all, of the pitfalls above.

\subsection{Excursion sets and excursion functions}

One solution to the problem of interpreting quantile and exceedance probability maps is to use methods that explicitly consider the \textit{joint} probability of events across prediction locations, as opposed to the \textit{marginal} probability at each location. The approach taken here requires that we decide, \emph{a priori}, on relevant values of the process and acceptable levels of uncertainty that are suitable given the context and aims of the analysis.  These values can then be used to construct summary maps that consider joint events across prediction locations.

We demonstrate this perspective using excursions sets and excursion functions \citep{bolin_excursion_2015}. By choosing acceptable levels of uncertainty, these methods avoid the issue of the default colour scale for the SD and CV maps which may potentially affecting our interpretation of results.  Excursion sets and excursion functions are also based on the joint probability of events across a set of locations.  For this reason they avoid the interpretability issues of the quantile maps.

We give the technical definition of excursion sets and functions below, using the same notation as \cite{bolin_excursion_2015}.  For readers unfamiliar with these methods, the notation can take some getting used to.  However, the precise mathematical definitions are useful when it comes to interpreting these methods.  We try to, as much as possible, give an informal explanation alongside the mathematical description.

The positive excursion set with level $u$ for a function $f(\bs)$ with domain $\Omega$ is $A_u^{+}(f) = \{ \bs \in \Omega ; f(\bs) > u \}$, i.e. the set of all locations in $\Omega$ where $f$ exceeds a threshold value $u$. For a random field, $\lambda(\bs)$, the positive excursion set with level $u$ and probability $1 - \alpha$ is
\begin{equation*}
E_{u,\alpha}^{+}(\lambda) = \argmax_{D}\{\lvert D \rvert : \mathbb{P}\left[D \subset A_u^{+}(\lambda)\right] \geq 1 - \alpha \} .
\end{equation*}
Note that $A_u^{+}(f)$ specifies a set of locations for which a function $f(\bs)$ exceeds a threshold value $u$ for \textit{every location} in the set. Therefore, the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is the \textit{largest} such set of locations for which \textit{realisations} of $\lambda(\bs)$ exceed the threshold $u$, \textit{simultaneously} for all locations in the set, with probability $1-\alpha$.  Note that we must decide on relevant choices for $u$ and $\alpha$ given the aims and context of the analysis.  Negative excursion sets are similarly defined by considering the probability of being below a threshold value.  Contour uncertainty can also be defined as the set difference between positive and negative excursion sets \citep{bolin_excursion_2015}.

Excursion sets can be estimated by considering candidate sets for $D$ of increasing size and a sequential integration scheme to estimate the required probabilities in a computationally efficient way.  A full description of the methods is given in \cite{bolin_excursion_2015} and a software implementation in the \texttt{excursions} package \citep{bolin_calculating_2018}.

\autoref{fig:excursions}A shows the positive excursion set with a level corresponding to 1 bird per hectare with probability 0.95 (hence $\alpha = 0.05$).  This figure can be interpreted in a natural way as the largest region for which the intensity is greater than 1 bird per hectare for every location within the region, with probability 0.95. 
\begin{figure}[!htb]
	\includegraphics{figures/excursions.pdf}
	\caption{A) The blue area is the positive excursion set with a level corresponding to 1 bird per hectare and probability 0.95.  B) The positive excursion function with a level corresponding to 1 bird per hectare.  The scale shows the probability values $1-\alpha$.}
	\label{fig:excursions}
\end{figure} 
This map could be used to define, for example, a `core region' for the \akepa{} population, with particular conservation importance.  Crucially, this map depends on a clear mathematical description of what a `core region' actually means in a statistical sense.  While any particular definition would be up for debate, the advantage is that the statistical methodology is taken as close as possible to the aims of the analysis.  Often a map of species density will be interpreted informally in many different ways (one of which is identifying hotspots for example).  Here we present an example of how a clear statistical definition can be used to produce a map that incorporates relevant definitions and uncertainties, and can be interpreted clearly.

To visualise how such excursion sets change with different levels of uncertainty we can use the excursion function $F_u^{+}(\bs) = \sup \{1 - \alpha ; \bs \in E_{u,\alpha}^+ \}$.  This defines, for each location, the largest possible probability $1 -\alpha$ for which that location would be in the excursion set defined using probability $1 - \alpha$.  I.e.  if we allow greater uncertainty, this function shows which locations would be included in the excursion set.

The excursion function with a level corresponding to 1 bird per hectare is shown in \autoref{fig:excursions}B.  It is clear from the figure that regions close to the edge of the excursion set with $\alpha = 0.05$ would be included if the $\alpha$ value were allowed to increase slightly.  Similarly, the core of the high density region seems to exceed 1 bird per hectare even for $\alpha$ levels close to zero.  There is a region to the north of the main population that would be included in the positive excursion set for 1 bird per hectare if the probability were allowed to be lower, around 0.5.  However, for regions in the north there is essentially no probability level for which those locations would be included.  This map could be used to identify, for example, secondary areas of potential importance beyond the core population that could be investigated further.

We chose the threshold of 1 bird per hectare and an error probability of 0.05 to demonstrate the approach.  However, multiple values for each of these can be considered and we imagine that, in most contexts, there would be value in considering various sets of thresholds and uncertainty levels.

@@

We have presented a new approach to analysing point transect distance sampling data in a Bayesian setting.  We chose to do this in a relatively simplified setting, considering a single year of data, with no covariate effects and using a GRF to explain spatial variation in density.  There are several natural extensions to the model.  Spatio-temporal models can also be fitted using \texttt{R-INLA}, which would allow multiple years of data to be modelled jointly, as was the case in \cite{camp_dsm_2020}.  Another extension is to develop the observation process model by considering other detection functions or including covariates, for example.  Factors such as weather conditions, observer expertise, animal behaviour and morphological traits are likely to affect detectability.  Future research should investigate the degree of complexity that can be well-approximated by the iterated INLA approach.

@@

\begin{itemize}
	\item \textbf{Selectivity analysis} In surveys of the marine environment, the size, length and weight of fish affects the likelihood of being caught in a trawl net.  The probability of capture is typically modelled as a logistic function with parameters that depend on morphological traits \citep{herrmann_understanding_2016, madsen_selectivity_2007, galbraith_demersal_1994}.  This function is analogous to the detection function in distance sampling models.  Our framework could allow, for example, the joint modelling of selectivity and spatial data.
	\item \textbf{Functional responses} The concept of a `functional response' captures the idea that the overall abundance of a resource or risk will affect a species' response to it \citep{holling_some_1959}.  These varying responses can be described mathematically using parametric equations that have the desired properties, e.g. an increasing function with a horizontal asymptote, to capture the notion of diminishing returns.  Such non-linear functions could be incorporated within our framework.  This flexibility in specifying model components could also be more widely applicable in areas such as public health and spatial epidemiology, where responses to conditions may suitably be modelled using non-linear relationships.
\end{itemize}
There are no doubt many more application areas that could make use of the additional flexibility that comes with allowing non-linear predictors.

>>>>

\autoref{fig:intensity-mean-cv}A shows a map of the posterior mean intensity field alongside the coefficient of variation (CV) (\autoref{fig:intensity-mean-cv}B) and standard deviation of the posterior intensity field (\autoref{fig:intensity-mean-cv}C) on a regular prediction grid.
\begin{figure}[!htb]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/intensity_mean_cv_sd.pdf}
		\caption{The A) mean, B) coefficient of variation and C) standard deviation of the posterior intensity field.  For the mean and standard deviaion the units are number of birds per m$^2$.}
		\label{fig:intensity-mean-cv}
	\end{center}
\end{figure}
This shows a region of higher intensity in the south and lower intensity in the north, agreeing with the broad patterns observed in a standard two-stage analysis of the same data \citep{camp_dsm_2020}.  The CV plot shows that the CV in the posterior intensity field is lower in areas with greater sampling effort and higher intensity. The standard deviation map shows stronger overall posterior variability in the south where intensity is larger.  Although survey effort is higher in the south, the increased uncertainty is likely due to the assumed log-normal relationship between the random field and the Poisson rate parameter.  For this reason we suggest using both the CV and SD to visualise uncertainty.

\autoref{fig:dist-mat-pairs} summarises the estimates of spatial structure in the model.  The posterior detection function (\autoref{fig:dist-mat-pairs}A) shows that the detection probability drops to around 0.25 at the maximum observable distance of 58 metres.
\begin{figure}[!htb]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/dist_mat_pairs.pdf}
		\caption{A) Posterior half-normal detection function. B) Posterior Mat\'ern correlation function for the GMRF spatial effect.  The black line is the posterior mean and the grey shaded area marks the 95\% credible interval. C) Boxplots of the frequency of pairwise distances based on 500 simulated datasets from the posterior predictive distribution. Red circles indicate the observed count of pairwise distances within each bin.}
		\label{fig:dist-mat-pairs}
	\end{center}
\end{figure}
The posterior Mat\'ern correlation function is shown in \autoref{fig:dist-mat-pairs}B, this represents the degree of spatial smoothing in the posterior intensity field. We note that this correlation function shows positive correlation on the log-scale. On the intensity scale the empirical range of correlation is shorter.  \autoref{fig:dist-mat-pairs}C summarises the model's ability to replicate the spatial structure in the intensity and detection process simultaneously by comparing the frequency of pairwise distances between all observations and those on data simulated from the joint posterior of the fitted model.  This is based on 500 Monte Carlo samples and is essentially an adjusted version of Ripley's $K$-function \citep{ripley_SecondorderAnalysisStationary_1976}, which accounts for the distance sampling observation process.  The model shows a good ability to reconstruct the spatial structure due to the distance to observe and spatially varying intensity surface.

To visualise the posterior abundance for the \akepa{} in Hakalau we use a Monte Carlo method based on sampling realisations of the posterior intensity field.  Let $N = \int_{\Omega}\lambda(\bs)\mathrm{d}\bs$, the integral of the intensity field over the study region.  A point estimate can be calculated by integrating the posterior mean intensity. To generate a full posterior estimate for abundance we approximate the posterior as $\pi(N | \bm{Y}) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda^{(i)}, \bm{Y})$ where $\lambda^{(1)}, \ldots, \lambda^{(m)}$ are samples from the posterior intensity field. Due to the properties of the log-Gaussian Cox process, each $\pi(N | \lambda = \lambda^{(i)}, \bm{Y})$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda^{(i)}(\bs)\mathrm{d}\bs$. 
\begin{figure}[!htb]
	\begin{center}
		\includegraphics{figures/N_posterior.pdf}
		\caption{Posterior abundance.  Grey area marks two Monte Carlo standard errors above and below the mean.}
		\label{fig:realized-abundance-posterior}
	\end{center}
\end{figure}
\autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.  This allows us to estimate the probability for any specific value of realised abundance $n$ which may be of interest to conservation managers.  This approach can also be used to generate credible intervals and other posterior summaries of interest.

Our abundance estimates cannot be directly compared to the results in \cite{camp_dsm_2020} since their two-stage model also includes a temporal correlation effect and we include an area of Hakalau that the authors in \cite{camp_dsm_2020} excluded.  Converting the above abundance estimates to density per hectare for the whole study region allows for some comparison with Figure 2 in \cite{camp_dsm_2020} which suggests that \akepa{} density was roughly between 0.5 and 1.5 birds per hectare in 2002 (considering both the design-based estimate and the spatio-temporal GAM estimate).  In our analysis we find a credible region of roughly 0.9 to 1.7 birds per hectare which overlaps significantly with the interval found in \cite{camp_dsm_2020}.

An attractive feature of the Bayesian one-stage model is that, by basing model outputs on posterior Monte Carlo samples, all the outputs incorporate the uncertainty in the detection model in a natural way.  There is no additional step required to incorporate detection function uncertainty beyond working with samples from the joint posterior.

@@

All figures that summarise the posterior intensity field have potential weaknesses that should be considered when communicating with stakeholders.  This sections discusses these weaknesses and presents and example of one approach that tries to address them.  The aim of this section is to show that our approach is readily extendible to produce bespoke model outputs.

Our presentation of the results above is broadly consistent with approaches taken in the species distribution modelling literature, although it is frustratingly common to see maps of point estimates without any accompanying communication of uncertainty.  The most common method to communicate uncertainty in maps of animal density is to produce maps of measures of predictive uncertainty, such as the standard deviation, CV \citep{fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014}, posterior quantiles or the probability of exceeding certain thresholds at each prediction location \citep{russell_avoidance_2016, wilson_hierarchical_2010}.  Whilst such maps are useful, they all mask certain properties of the intensity field that may be important when communicating the results of the analysis. For example, the posterior mean will always be smoother than realisations of the field.  If there is interest in understanding patterns of spatial clustering, the posterior mean map presents a misleadingly homogeneous picture.  This highlights the fact that even a relatively innocuous map of the posterior mean can mask certain features of the random field. \autoref{fig:intensity-realizations} shows three realisations of the posterior field, all of which show more heterogeneity in the high intensity region compared to the posterior mean.
\begin{figure}[!htb]
	\includegraphics{figures/intensity_realized.pdf}
	\caption{Three realizations of the posterior intensity field}
	\label{fig:intensity-realizations}
\end{figure}

The CV map (\autoref{fig:intensity-mean-cv}B) is also intended to communicate uncertainty.  However, CV values will generally be higher in regions of low predicted intensity, particularly if the posterior standard deviation is relatively consistent across the study region or positively correlated with the intensity.  Higher CV values in Figure \ref{fig:intensity-mean-cv}B represent areas of lower intensity but not necessarily higher uncertainty.  However, there is also variable sampling effort within the study region, with higher CV values in the region with lower effort.  It is not clear what impact this has on the CV values and this makes the map hard to interpret, even for statisticians who understand the model well. The standard deviation (SD) map (\autoref{fig:intensity-mean-cv}C) shows spatially varying SD values also intended to indicate uncertainty.  However, a default colour scale will nearly always show some spatial variation in uncertainty, with some regions of relatively high and low uncertainty.  Whether these relative differences matter is dependent on the overall context of the analysis.  Some levels of variability may be acceptable in some contexts and unacceptable in others.

Quantile maps (Figure \ref{fig:intensity-quantiles}) are also difficult to interpret. The temptation is to perceive these maps as showing possible intensity surfaces that could have produced the observed data.  Stake holders may interpret the 0.025 quantile map as a kind of lower bound intensity surfaces and the 0.975 quantile map as an upper bound. However, presenting these quantiles together in a single map obscures the fact that it is vanishingly unlikely for the random field to \textit{simultaneously} achieve the 0.025 or 0.975 quantiles at all prediction locations.  A similar problem holds for maps of the marginal probability of exceeding a threshold at each prediction location. 
\begin{figure}[!htb]
	\begin{center}
		\includegraphics{figures/intensity_quantiles.pdf}
		\caption{Pointwise posterior quantiles for the intensity:  A) 0.025 quantile B) 0.975 quantile}
		\label{fig:intensity-quantiles}
	\end{center}
\end{figure}
These caveats make the map difficult to use for most non-statistically trained audiences and even trained statisticians may misinterpret these quantile maps.  To demonstrate the misleading nature of these maps, if we (incorrectly) treat the lower and upper quantile plots as though they are intensity surfaces and integrate them, then we obtain an abundance estimate of approximately 2,900 for the 0.025 quantile map and 10,700 for the 0.975 quantile map.  However, these abundance estimates are not supported by the posterior for abundance (\autoref{fig:realized-abundance-posterior}).

Broadly speaking, we advocate for moving beyond maps of uncertainty presented for their own sake and towards model outputs that directly relate to the wider objectives of the analysis. We conclude that, although all useful in their own way, each of the mapped summaries of the posterior intensity field suffer from some weaknesses, whether through masking certain properties of the random field or through being difficult or impossible to interpret. In the following section we suggest an approach that avoids some, but not all, of the pitfalls above.

\subsection{Identification of hotspots}

In this section we consider the hypothetical challenge of using our model to identify population hotspots within Hakalau refuge.  This acts as a focus for communication that directs our choices when producing model outputs.

A solution to the problem of interpreting quantile and exceedance probability maps is to use methods that explicitly consider the joint probability of events across all prediction locations. One option is to use excursions sets and excursion functions \citep{bolin_excursion_2015}.  This requires users to choose thresholds of interest (e.g. the definition of a hotspot), and to select an acceptable level of uncertainty in knowing whether the field exceeds this threshold.  This forces us to consider this explicitly and leaves us less at the whim of an arbitrary colour scale.  Crucially, excursion sets and functions can be estimated using Monte Carlo samples and so fit well within our one-stage Bayesian approach.

We provide the technical definition of excursion sets and functions below, using the same notation as \cite{bolin_excursion_2015}.  For readers unfamiliar with these methods, the notation can take some getting used to.  However, the precise mathematical definitions are useful when it comes to interpreting these methods.  We try to, as much as possible, give an informal explanation alongside the mathematical description.  The positive excursion set with level $u$ for a deterministic function $f(\bs)$ with domain $\Omega$ is $A_u^{+}(f) = \{ \bs \in \Omega ; f(\bs) > u \}$, i.e. the set of all locations in $\Omega$ where $f$ exceeds a threshold value $u$. For a random field, $\lambda(\bs)$, the positive excursion set with level $u$ and probability $1 - \alpha$ is
\begin{equation*}
E_{u,\alpha}^{+}(\lambda) = \argmax_{D}\{\lvert D \rvert : \mathbb{P}\left[D \subset A_u^{+}(\lambda)\right] \geq 1 - \alpha \} .
\end{equation*}
In other words, the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is the \textit{largest} set of locations for which realisations of $\lambda(\bs)$ exceed the threshold $u$, \textit{simultaneously}, for all locations in the set, with probability $1-\alpha$.  Note that we must decide on relevant choices for $u$ and $\alpha$ given the aims and context of the analysis.  Negative excursion sets are similarly defined.  Excursion sets can be estimated by considering candidate sets for $D$ of increasing size and a sequential integration scheme to estimate the required probabilities in a computationally efficient way.  A full description of the methods is given in \cite{bolin_excursion_2015} and a software implementation in the \texttt{excursions} package \citep{bolin_calculating_2018}.

\autoref{fig:excursions}A shows the positive excursion set with a level corresponding to 1 bird per hectare with probability 0.95 (hence $\alpha = 0.05$).  Note that we use this threshold as the definition of a hotspot, and proceed with generating model results that incorporate this definition.  This figure can be interpreted in a natural way as the largest region for which the intensity is greater than 1 bird per hectare for every location within the region, with probability 0.95. 
\begin{figure}[!htb]
	\includegraphics{figures/excursions.pdf}
	\caption{A) The blue area is the positive excursion set with a level corresponding to 1 bird per hectare and probability 0.95.  B) The positive excursion function with a level corresponding to 1 bird per hectare.  The scale shows the probability values $1-\alpha$.}
	\label{fig:excursions}
\end{figure} 
This map could be used to define, for example, a `core region' for the \akepa{} population, with particular conservation importance.  Crucially, this map depends on a clear mathematical description of what a `core region' actually means in a statistical sense.  While any particular definition would be up for debate, the advantage is that the statistical methodology is taken as close as possible to the aims of the analysis.  Often a map of species density will be interpreted informally in many different ways (one of which is identifying hotspots).  To visualise how such excursion sets change with different levels of uncertainty we can use the excursion function $F_u^{+}(\bs) = \sup \{1 - \alpha ; \bs \in E_{u,\alpha}^+ \}$.  This defines, for each location, the largest possible probability $1 -\alpha$ for which that location would be in the excursion set defined using probability $1 - \alpha$.  I.e.  if we allow greater uncertainty, this function shows which locations would be included in the excursion set.

The excursion function with a level corresponding to 1 bird per hectare is shown in \autoref{fig:excursions}B.  It is clear from the figure that regions close to the edge of the excursion set with $\alpha = 0.05$ would be included if the $\alpha$ value were allowed to increase slightly.  Similarly, the core of the high density region seems to exceed 1 bird per hectare even for $\alpha$ levels close to zero. This map could be used to identify, for example, secondary areas of potential importance beyond the core population that could be investigated further.

@@

There are several natural extensions to the model presented here.  Spatio-temporal models can also be fitted using \texttt{R-INLA}, which would allow multiple years of data to be modelled jointly, as was the case in \cite{camp_dsm_2020}.  Another extension is to develop the observation process model by considering other detection functions or including covariates, for example.  Factors such as weather conditions, observer expertise, animal behaviour and morphological traits are likely to affect detectability.

@@

\begin{itemize}
	\item \textbf{Selectivity analysis} In surveys of the marine environment, the size, length and weight of fish affects the likelihood of being caught in a trawl net.  The probability of capture is typically modelled as a logistic function with parameters that depend on morphological traits \citep{herrmann_understanding_2016, madsen_selectivity_2007, galbraith_demersal_1994}.  This function is analogous to the detection function in distance sampling models.  Our framework allows for the joint modelling of selectivity and spatial data.
	\item \textbf{Functional responses} The concept of a `functional response' captures the idea that the overall abundance of a resource or risk will affect a species' response to it \citep{holling_some_1959}.  These varying responses can be described mathematically using parametric equations that have the desired properties, e.g. an increasing function with a horizontal asymptote, to capture the notion of diminishing returns.  Such non-linear functions could be incorporated within our framework
\end{itemize}

The popularity of using INLA for spatial modelling has grown, in large part, due to the computational efficiency of the approach.  However, this efficiency comes at the cost of reduced flexibility in model specification over, say, a more general probabilistic programming languages such as \texttt{JAGS} \citep{plummer_jags_2017} or \texttt{Stan} \citep{stan_Rstan_2020}.  The iterated INLA approach allows a wider class of models to be considered whilst retaining much of the computational benefits of INLA.  In addition to computational efficiency, there are some features of INLA that are not generally available in other packages, such as: (i) the ability to model using GRFs directly on the sphere, avoiding projecting to $\mathbb{R}^2$ \citep{lindgren_explicit_2011}; (ii) GRFs that account for complex barriers such as coastlines and islands, to avoid inappropriately smoothing across these \citep{bakka_NonstationaryGaussianModels_2019}; and (iii) support for joint likelihood models, allowing modelling of, for example, marked point processes \citep{illian_FittingComplexEcological_2013}, and combining multiple data sources.  These approaches can now readily be incorporated into distance sampling analyses, as well as broader spatial ecology and spatial statistics applications.
